---
title: "Estimated-Phytoplankton-Size"
author: "phoebe.woodworth-jefcoats@noaa.gov"
format: 
  docx:
    reference-doc: Palau-Reference-Doc.docx
---

```{r}
#| include: false
### Load libraries
library(tidyverse)
library(lubridate)
library(here)
library(stringr)
library(nmfspalette)
library(rerddap)
```

```{r}
#| include: false
# Set report year (RptYr), to make things easier
RptYr <- 2022

# Set path to variable: Sea_Surface_Temperature
# This is where the data are and where the plots will go
Dir <- here("Estimated_Phytoplankton_Size")
```

```{r}
#| include: false
### Load data
# Thanks to Melanie Abecassis and the OceanWatch training she led, whose code I'm borrowing here

# Bounding box, from John Marra via email:
lon_range <- c(129.4088, 137.0541)
lat_range <- c(1.5214, 11.6587)

# OceanWatch ERDDAP node
ERDDAP_Node <- "https://oceanwatch.pifsc.noaa.gov/erddap/"

# The variable name and other information can be found at the site above.
# I'm omitting the sleuthing from the code for the sake of brevity and
# because I'm familiar with the data.

# Access the data and all its attributes (this takes a bit)
md50_all <- griddap(url = ERDDAP_Node, 'md50_exp', 
               time = c('1998-01-01', paste(RptYr, '-12-01', sep = "")),
               latitude = lat_range, longitude = lon_range,
               fields = "MD50")
# Get just the chlorophyll data and its spatiotemporal coordinates
md50 <- md50_all$data
```

```{r}
#| include: false
# Monthly spatial average
md50_ts <- md50 |>
  group_by(time) |>
  summarise(MD50 = mean(MD50, na.rm = TRUE))
```

```{r}
#| include: false
### Linear fit
n_obs <- seq(1, length(md50_ts$MD50), 1)
md50_lm <- lm(md50_ts$MD50 ~ n_obs)

# summary(md50_lm) shows that there's a weakly significant 
# trend over time for this indicator.  There are some automated
# checks that can be added to make sure this is still the case
# in future years.

# Change over time
delta_md50 <- md50_lm$fitted.values[length(n_obs)] - md50_lm$fitted.value[1]
```

## __Indicator: Estimated Phytoplankton Size__
Phytoplankton size provides insight into ecosystem productivity with larger 
phytoplankton generally supporting more productive ecosystems with larger fish. 
Some climate change projections suggest a shift towards smaller phytoplankton, 
particularly in the ocean’s oligotrophic gyres, potentially reducing food available 
to all trophic levels. Phytoplankton size varies greatly across the Pacific basin, 
with larger phytoplankton generally found at higher latitudes and closer to coastlines.  
Phytoplankton size also varies with natural climate cycles such as ENSO, with larger 
phytoplankton more prevalent across the equatorial Pacific during cooler La Niña 
periods and vice versa during El Niño.

Estimated median phytoplankton size can be derived from satellite remotely sensed sea 
surface temperature and chlorophyll-a concentration.  The basin-wide average (1998 – `r RptYr - 1`) 
is shown below.  The Palau EEZ is also highlighted as a spatial average (1998 – `r RptYr - 1`)
and a time series averaged over this area.  A statistically significant (p < 0.05) decline 
in estimated median phytoplankton size was detected over this area.  From 1998 through `r RptYr`,
estimated median phytoplankton size declined by `r signif(abs(delta_md50),2)` $\mu$m Equivalent 
Spherical Diameter (ESD).

```{r}
#| echo: false
# Note that this needs to be 'echo' and not 'include' so the figure will show up

### Plot the time series
# Create axes limits to make things simpler
# These were determined through looking at quick rough plots and data limits
md50_xlim <- c(min(ymd_hms(md50_ts$time)), max(ymd_hms(md50_ts$time)))
md50_ylim <- c(0.8, 1.3)

# Access the NMFS color palette
oceans <- nmfs_palette("oceans")(3)
crustacean <- nmfs_palette("crustacean")(4)

# Plot
plot(ymd_hms(md50_ts$time), md50_ts$MD50, type = "l", lwd = 2, col = oceans[2], 
     xlim = md50_xlim, ylim = md50_ylim, xlab = " ", ylab = "Estimated Phytoplankton Size (ESD)",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
par(new = TRUE)
plot(ymd_hms(md50_ts$time), md50_lm$fitted.values, type = "l", lwd = 2, col = crustacean[1], 
     xlim = md50_xlim, ylim = md50_ylim, xlab = " ", ylab = " ",
     xaxt = "n", yaxt = "n", xaxs = "i", yaxs = "i")
axis((1), at = ymd_hms(md50_ts$time[seq(1, length(n_obs), 12)]), tck = 0.025, labels = year(make_date(seq(1998, RptYr, 1))))
axis((2), at = seq(0.8, 1.3, 0.1), tck = 0.025, las = 1)
axis((3), at = ymd_hms(md50_ts$time[seq(1, length(n_obs), 12)]), tck = 0.025, labels = FALSE)
axis((4), at = seq(0.8, 1.3, 0.1), tck = 0.025, labels = FALSE)
# _axt = "n" removes tick labels so that they can be customized later 
# _axs = "i" removes whitespace beyond axes maxima
```
  
